A free/open-source machine translation system for English to Kazakh

### 2015-03-05 engkaz-eamt-2015/engkaz-eamt-2015.tex
added kazakh-english paper, answered to some comments
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@59215 72bbbca6-d526-0410-a7d9-f06f51895060

% Authors:
% Aida, Mikel, Fran, Inari, ... ?

@@

We set out to build a rule-based machine translation system, because for statistical machine translation 
system we need large sentence-aligned parallel corpora, which are not freely available
for the English--Kazakh pair. An English--Kazakh statistical MT system has however recently been built by Google,\footnote{\url{http://translate.google.com}}; 
other work in this language pair was performed by building a small parallel corpus and using Moses \citep{assylbekov14}. Existing commercial systems for English to Kazakh (Sanasoft\footnote{\url{http://www.sanasoft.kz/c/ru/node/47}} and Trident\footnote{\url{http://www.translate.ua/us/on-line}}) both appear to be rule-based. 
Rule-based machine translation, in which developers 
write up dictionaries and grammatical rules that are applied by an engine, in some 
unusual case will produce better translation.
\todo{What does the last sentence mean? Clarify!}
This paper describes work in progress of development of machine translation system 
for English--Kazakh using the Apertium free/open-source machine 
translation platform \citep{forcada11}. %(http://www.apertium.org).

@@

\item[Post-generator:] Performs some minor orthographical operations in the target language. This module is not used in the English--Kazakh system.\todo{Or is it?}

@@

The Kazakh dictionary is turned into a
morphological transducer, using the Helsinki Finite
State Toolkit \citep{hfst/2011}. It uses the \texttt{lexc} formalism
for defining lexicons through word classes and subclasses, and the
\texttt{twol} formalism for morphophonological rules such as vowel
harmony, desonorization, nasalization, etc.  Depending on the
preceding vowels and consonants, the plural suffix \emph{-LAr}\footnote{Uppercase Latin letters are used for archiphonemes (actually archigraphemes) that are realised as phonemes (actually graphemes) after morphophonological rules have been applied} could
become \emph{-тар}, \emph{-тер}, \emph{-lar}, \emph{-ler}, \emph{-dar}, \emph{-ler} depending on vowel harmony: kitap+tar,
mektep+ter, etc. \todo{Aida, please typeset this example in Cyrillic; braces are not used here, but are used later; homogeneize notation in paper.}

@@

The English--Kazakh bilingual dictionary provides correspondences between English LFs and Kazakh LFs. Ambiguity is allowed: one of the LFs will be chosen by 
lexical-selection rules depending on context. This dictionary currently contains 13,130 stems \citep{sundetova13a}.\todo{later you say 13,135}

@@

Lexical-selection rules choose one of the alternative target-language LFs corresponding to one source-language LF, as described in \cite{tyers12a}. Alternative translations are defined in bilingual dictionary by 
multiple entries for each source-language LF. For example, the adjective \emph{beautiful} could be translated as \emph{сұлу}, if 
the following noun is a person: \emph{beautiful girl} \(\to\) \emph{сұлу қыз}; or as \emph{әдемі} or  \emph{көркем}, if the following noun means 
place or \todo{or what?} \emph{beautiful mountains} \(\to\) \emph{әдемі таулар}.  Table~\ref{table:lexsel} shows some 
lexical-selection rules:\todo{but this rule is not in the table!!!}
\begin{table*}
 \centering
 \begin{tabular}{|l|l|l|l|}
    \hline
    \textbf{SL word \(w\)} & \textbf{TL word} & \textbf{Context} & \textbf{Example} \\
    \hline
    \multirow{4}{*}{residence} & \multirow{2}{*}{мекен}      & \multirow{2}{*}{default} & I live in a residence. \\
                               &                             &                     & (Мен) мекенде өмір сүремін. \\
                               & \multirow{2}{*}{резиденция} & \multirow{2}{*}{\(w\)\_ of \_ president} & I see the residence of the president \\ 
                               &                             &                                     & (Мен) президенттің резиденциясын көремін \\
    \hline
    \multirow{4}{*}{boot} & \multirow{2}{*}{бәтеңке} & default & I bought boots. \\
                          &                          &     &  Мен бәтеңкелерді сатып алдым \\

@@

    \hline
 \end{tabular}
  \caption{Example lexical-selection rules}
  \label{table:lexsel}
\end{table*}
\todo{I do not understand the \emph{I see anything} example. Does it mean \emph{anything} as \emph{whatever}? I do not know how to put the \(w\) in this example}@

@@

This section describes the current structural transfer in \texttt{apertium-eng-kaz} (revision ..., ).\todo{Complete! 58998? Move to results?} English--Kazakh 
chunker rules, interchunk rules and an additional clean-up stage will be described in detail in the following 3 sections.

@@

The second round of structural transfer is currently performed by a %proof-of-concept 
set of 90 rules, representative of following operations:\todo{I have removed \emph{proof-of-concept} here}
\begin{itemize}
\item Inter-chunk agreement between subject noun phrase and verb phrase in their person and number.
\item Assigning case to noun phrases (which are generated without case by the chunker): for instance, 
  accusative case for objects (\emph{I bought the table} $\rightarrow$ \emph{Мен үстелді  сатып алдым}), genitive case for 
  obligatory constructs (\emph{I must see} $\rightarrow$ \emph{менің көруім керек}),  dative case for the verb \emph{to 
  need} (\emph{I need a doctor} $\rightarrow$ \emph{Маған дәрігер керек}), locative case for 
  possession (\emph{I have a book} $\rightarrow$ \emph{Менде кітап бар}), etc.\ 
\item Reordering: placing of object before verb ( I[1] bought[2] the table[3] $\rightarrow$ Мен[1] үстелді[3] сатып алдым[2]), \todo{italics for examples please!}
  placing of prepositional pharses before the verb  (They[1] played[2] on top of the tree [3] $\rightarrow$ Олар[1] ағаштың үстінде[3] ойнады [2]), etc.\ 
\item Adding question word ма/ме/ба/бе in the end of question, if the question mark chunk ``?'' is detected (Did<VP\_ques> you<NP> watch<VP> last film<NP> ?<Q\_m> $\rightarrow$ Сіз<NP> соңғы фильмді<NP> көрдіңіз<VP> бе<ques> ?<Q\_m>).\todo{Notation: use brackets as above!}
\item Place possessive for long noun + noun + noun structures(The university of city of Kazakhstan -- Қазақстан қаласының университеті).\todo{Explain better, as this is a rule to avoid repeating -NIng, -NIng, etc. All possessives are in place}
\end{itemize}

@@

\paragraph{Postchunk:}\todo{Nothing is said about postchunk --- this talks about the additional cleaning stage}
An additional cleaning stage was created for removing extra tags which were ot determined during the chunker and 
interchunk steps, for instance, a noun was not determined to be acting as an object and therefore did not get accusative case; such a noun would have been assigned the  \texttt{<CD>} (``case to be determined'') tag, which after ``cleaning'' would be set to nominative. Other operations  carried out at this stage ensure
the agreement between noun or adjective and the copula verb ``е''(\emph{Мен дәрігер+е}<p1 person singular>), \todo{add result for copula} or deciding the actual form of the question particle according to the preceding word (\emph{Сіз келдіңіз} + \emph{ме}? \(\to\) \emph{Сіз келдіңіз} + \emph{бе}?).

@@

Current system can translate simple sentences and questions. The English--Kazakh bilingual dictionary 
contains 13,135 entries, \todo{13130 before!} There are 168 ``chunker'' rules and 148 interchunk rules. 
Evaluating was tested against revision 58998 in the Apertium Subversion repository. Lexical coverage is calculated 
over EuroParl,\footnote{\url{http://www.statmt.org/europarl/v7/es-en.tgz}} SETimes,\footnote{\url{http://nlp.ffzg.hr/data/corpora/setimes/setimes.en-tr.txt.tgz}} NewsCommentary,\footnote{\url{http://www.statmt.org/wmt13/training-parallel-nc-v8.tgz}} Wikipedia.\footnote{\url{http://dumps.wikimedia.org/enwiki/20140402/enwiki-20140402-pages-articles.xml.bz2}; last checked 04/06/2014} 
Table~\ref{table:coverage} presents the size of the corpora and the vocabulary coverage of the system for the given corpus.

@@

As can be seen,\todo{where!} the coverage of dictionary is good, average score is 94\%. Our system outperforms the other 
available RBMT system, but fall short of the state-of-the-art performance.\todo{represented by...} However, unlike the state of 
the art, which uses unavailable corpora, our system is free and open and can be improved.

@@

We have presented the design of a free/open-source rule-based MT system from English to Kazakh. The current English--Kazakh machine translation 
system already successfully translates noun-phrases, verb-phrases, prepositional-phrases, and adjectival-phrases, and 
contains a good vocabulary for testing purposes. 
We plan to continue developing the English--Kazakh pair, going to extend the coverage to 98\% on the corpora, 
which we used before. Additionally, we will improve the quality of translation by adding more rules, such like 
constraint grammar rules, structural transfer and lexical rules. The future plan is to use the created data with other 
open-source MT system with Turkic languages or system with Kazakh language on right side to make transfer systems 
between the Turkic or other language pairs. Related work is currently ongoing with Russian--Kazakh.\todo{Say something about Kazakh--English.}
Our system is available as free/open-source software and the whole system may be downloaded 
from SourceForge.\footnote{\url{https://svn.code.sf.net/p/apertium/svn/incubator/apertium-eng-kaz}}

>>>>

% Authors:
% Aida, Mikel, Fran, Inari, ... ?
%Aida Sundetova
%Al-Farabi Kazakh National University, Al-Farabi av., 71, 050040 Almaty, Kazakhstan

@@

We set out to build a rule-based machine translation system, because for statistical machine translation 
system we need large sentence-aligned parallel corpora, which are not freely available
for the English--Kazakh pair. An English--Kazakh statistical MT system has however recently been built by Google,\footnote{\url{http://translate.google.com}}; 
other work in this language pair was performed by building a small parallel corpus and using Moses \citep{assylbekov14}. Existing commercial systems for English to Kazakh (Sanasoft\footnote{\url{http://www.sanasoft.kz/c/ru/node/47}} and Trident\footnote{\url{http://www.translate.ua/us/on-line}}) both appear to be rule-based. 
Rule-based machine translation, in which developers 
write up dictionaries and grammatical rules that are applied by an engine, in some 
case, where statistics not work well, will produce better translation.
\todo{What does the last sentence mean? Clarify!}
This paper describes work in progress of development of machine translation system 
for English--Kazakh using the Apertium free/open-source machine 
translation platform \citep{forcada11}. %(http://www.apertium.org).

@@

\item[Post-generator:] Performs some minor orthographical operations in the target language. 

@@

The Kazakh dictionary is turned into a
morphological transducer, using the Helsinki Finite
State Toolkit \citep{hfst/2011}. It uses the \texttt{lexc} formalism
for defining lexicons through word classes and subclasses, and the
\texttt{twol} formalism for morphophonological rules such as vowel
harmony, desonorization, nasalization, etc.  Depending on the
preceding vowels and consonants, the plural suffix \emph{-LAr}\footnote{Uppercase Latin letters are used for archiphonemes (actually archigraphemes) that are realised as phonemes (actually graphemes) after morphophonological rules have been applied} could
become \emph{-тар}, \emph{-тер}, \emph{-лар}, \emph{-лер}, \emph{-дар}, \emph{-лер} depending on vowel harmony: кітап+тар,
мектеп+тер, etc. \todo{Aida, please typeset this example in Cyrillic; braces are not used here, but are used later; DONE  homogeneize notation in paper.}

@@

The English--Kazakh bilingual dictionary provides correspondences between English LFs and Kazakh LFs. Ambiguity is allowed: one of the LFs will be chosen by 
lexical-selection rules depending on context. This dictionary currently contains 13,135 stems \citep{sundetova13a}.\todo{later you say 13,135}

@@

Lexical-selection rules choose one of the alternative target-language LFs corresponding to one source-language LF, as described in \cite{tyers12a}. Alternative translations are defined in bilingual dictionary by 
multiple entries for each source-language LF. For example, the adjective \emph{beautiful} could be translated as \emph{сұлу}, if 
the following noun is a person: \emph{beautiful girl} \(\to\) \emph{сұлу қыз}; or as \emph{әдемі} or  \emph{көркем}, if the following noun means 
place or \todo{or what? - mountains} \emph{beautiful mountains} \(\to\) \emph{әдемі таулар}.  Table~\ref{table:lexsel} shows some 
lexical-selection rules:\todo{but this rule is not in the table!!! - should I add it or just remove upper example??}
\begin{table*}
 \centering
 \begin{tabular}{|l|l|l|l|}
    \hline
    \textbf{SL word \(w\)} & \textbf{TL word} & \textbf{Context} & \textbf{Example} \\
    \hline
    \multirow{4}{*}{residence} & \multirow{2}{*}{мекен}      & \multirow{2}{*}{default} & I live in a residence. \\
                               &                             &                     & (Мен) мекенде өмір сүремін. \\
                               & \multirow{2}{*}{резиденция} & \multirow{2}{*}{\(w\)\_ of \_ president} & I see the residence of the president \\ 
                               &                             &                                     & (Мен) президенттің резиденциясын көремін \\
    \hline
    \multirow{4}{*}{boot} & \multirow{2}{*}{бәтеңке} & default & I bought boots. \\
                          &                          &     &  Мен бәтеңкелерді сатып алдым \\

@@

    \hline
 \end{tabular}
  \caption{Example lexical-selection rules}
  \label{table:lexsel}
\end{table*}
\todo{Yes, it is mean I see whatever - I do not understand the \emph{I see anything} example. Does it mean \emph{anything} as \emph{whatever}? I do not know how to put the \(w\) in this example}@

@@

This section describes the current structural transfer in \texttt{apertium-eng-kaz} (revision 58998, ).\todo{Complete! 58998? Move to results?} English--Kazakh 
chunker rules, interchunk rules and an additional clean-up stage will be described in detail in the following 3 sections.

@@

The second round of structural transfer is currently performed by a %proof-of-concept 
set of 90 rules, representative of following operations:\todo{I have removed \emph{proof-of-concept} here}
\begin{itemize}
\item Inter-chunk agreement between subject noun phrase and verb phrase in their person and number.
\item Assigning case to noun phrases (which are generated without case by the chunker): for instance, 
  accusative case for objects (\emph{I bought the table} $\rightarrow$ \emph{Мен үстелді  сатып алдым}), genitive case for 
  obligatory constructs (\emph{I must see} $\rightarrow$ \emph{менің көруім керек}),  dative case for the verb \emph{to 
  need} (\emph{I need a doctor} $\rightarrow$ \emph{Маған дәрігер керек}), locative case for 
  possession (\emph{I have a book} $\rightarrow$ \emph{Менде кітап бар}), etc.\ 
\item Reordering: placing of object before verb ( I[1] bought[2] the table[3] $\rightarrow$ Мен[1] үстелді[3] сатып алдым[2]), \todo{italics for examples please!}
  placing of prepositional pharses before the verb  (They[1] played[2] on top of the tree [3] $\rightarrow$ Олар[1] ағаштың үстінде[3] ойнады [2]), etc.\ 
\item Adding question word ма/ме/ба/бе in the end of question, if the question mark chunk ``?'' is detected (Did[VP\_ques] you[NP] watch[VP] last film[NP] ?[Q\_m] $\rightarrow$ Сіз[NP] соңғы фильмді[NP] көрдіңіз[VP] бе<ques> ?[Q\_m]).\todo{Notation: use brackets as above!}
\item Place possessive for long noun + noun + noun structures(The university of city of Kazakhstan -- Қазақстан қаласының университеті).\todo{Explain better, as this is a rule to avoid repeating -NIng, -NIng, etc. All possessives are in place}
\end{itemize}

@@

\paragraph{Postchunk and cleanup:}\todo{Can I do that?  --Nothing is said about postchunk --- this talks about the additional cleaning stage}
An additional cleaning stage was created for removing extra tags which were ot determined during the chunker and 
interchunk steps, for instance, a noun was not determined to be acting as an object and therefore did not get accusative case; such a noun would have been assigned the  \texttt{<CD>} (``case to be determined'') tag, which after ``cleaning'' would be set to nominative. Other operations  carried out at this stage ensure
the agreement between noun or adjective and the copula verb ``е''(\emph{Мен дәрігер+е}<p1 person singular>), \todo{add result for copula} or deciding the actual form of the question particle according to the preceding word (\emph{Сіз келдіңіз} + \emph{ме}? \(\to\) \emph{Сіз келдіңіз} + \emph{бе}?).

@@

Current system can translate simple sentences and questions. The English--Kazakh bilingual dictionary 
contains 13,135 entries, \todo{13130 before!-changed} There are 168 ``chunker'' rules and 148 interchunk rules. 
Evaluating was tested against revision 58998 in the Apertium Subversion repository. Lexical coverage is calculated 
over EuroParl,\footnote{\url{http://www.statmt.org/europarl/v7/es-en.tgz}} SETimes,\footnote{\url{http://nlp.ffzg.hr/data/corpora/setimes/setimes.en-tr.txt.tgz}} NewsCommentary,\footnote{\url{http://www.statmt.org/wmt13/training-parallel-nc-v8.tgz}} Wikipedia.\footnote{\url{http://dumps.wikimedia.org/enwiki/20140402/enwiki-20140402-pages-articles.xml.bz2}; last checked 04/06/2014} 
Table~\ref{table:coverage} presents the size of the corpora and the vocabulary coverage of the system for the given corpus.

@@

As can be seen from table of coverage,\todo{where!-table of coverage} the coverage of dictionary is good, average score is 94\%. Our system outperforms the other 
available RBMT system, but fall short of the state-of-the-art performance.\todo{represented by...} However, unlike the state of 
the art, which uses unavailable corpora, our system is free and open and can be improved.

@@

We have presented the design of a free/open-source rule-based MT system from English to Kazakh. The current English--Kazakh machine translation 
system already successfully translates noun-phrases, verb-phrases, prepositional-phrases, and adjectival-phrases, and 
contains a good vocabulary for testing purposes. 
We plan to continue developing the English--Kazakh pair, going to extend the coverage to 98\% on the corpora, 
which we used before. Additionally, we will improve the quality of translation by adding more rules, such like 
constraint grammar rules, structural transfer and lexical rules. The future plan is to use the created data with other 
open-source MT system with Turkic languages or system with Kazakh language on right side to make transfer systems 
between the Turkic or other language pairs. Related work is currently ongoing with Russian--Kazakh and Kazakh--English\citep{sundetova14}.\todo{Say something about Kazakh--English.}
Our system is available as free/open-source software and the whole system may be downloaded 
from SourceForge.\footnote{\url{https://svn.code.sf.net/p/apertium/svn/incubator/apertium-eng-kaz}}
